Playing Atari Breakout with Keras and R

### 2018-07-13 article/article.Rmd
continuing the article

---
title: "Playing Atari Breakout with Keras and R"
description: |
  In this article we will implement the famous paper from DeepMind: *Playing Atari 
  with Deep Reinforcement Learning* in R using Keras. This paper introduced the 
  first deep learning model that successfully learned control policies directly 
  from high-dimensional sensory input using reinforcement learning.
author:
  - name: Daniel Falbel 
    url: http://github.com/dfalbel
    affiliation: Curso-R
    affiliation_url: http://curso-r.com
  - name: Fernando Barscevicius
    url: 
    affiliation: Curso-R
    affiliation_url: http://curso-r.com
date: 07-17-2018
creative_commons: CC BY
repository_url: https://github.com/dfalbel/rl-keras
output: 
  radix::radix_article:
    self_contained: false
---

@@

Now that we have Gym installed we can "make" an environment and run some random actions to understand how it works.

@@

Gym does not document which action is performed by each integer value, but we can figure out by running actions in the emulator and looking at the rendered output. We can also get the number of possible actions by accessing the `env$action_space$n` member.

In our particular case, ie. the Atari Breakout game, there are 4 possible actions and the following table maps actions to corresponding integer values.

|

>>>>

---
title: "Playing Atari Breakout with Keras and R"
description: |
  In this article we will implement the famous paper from DeepMind: *Playing Atari 
  with Deep Reinforcement Learning* in R using Keras. This paper introduced the 
  first deep learning model that successfully learned control policies directly 
  from high-dimensional sensory input using reinforcement learning.
author:
  - name: Daniel Falbel 
    url: http://github.com/dfalbel
    affiliation: Curso-R
    affiliation_url: http://curso-r.com
  - name: Fernando Barscevicius
    url: 
    affiliation: Marketdata
    affiliation_url: http://marketdata.com.br
date: 07-17-2018
creative_commons: CC BY
repository_url: https://github.com/dfalbel/rl-keras
output: 
  radix::radix_article:
    self_contained: false
bibliography: biblio.bib
---

@@

Now that we have Gym installed we can "make" an environment and run some random actions to understand how it works.
We are going to use an environment called BreakoutDeterministic-v4. It differs from the canonical Breakout-v0 in two ways related noise added when performing actions:

* It doesn't repeat the performed action for more randomly chosen number of frames.
* It doesn't randomly repeat the previous action.

These differences make BreakoutDeterministic-v4 easier to learn than Breakout-v0.

@@

The output of running a `step` in the environment is the next state of the game. The sate is represented by a list with the following structure:

```
List of 4
 $ : int [1:210, 1:160, 1:3] 0 0 0 0 0 0 0 0 0 0 ...
 $ : num 0
 $ : logi FALSE
 $ :List of 1
  ..$ ale.lives: int 5
```

The first element is an RGB array of the next frame, the second element is the reward received by performing the action and the thir is a boolean value indicating if the game has ended. The fourth element can contain other informations about the game, in this case, it returns the number of lives.

We can get the meaning of each action with `env.unwrapped.get_action_meanings()`. In our particular case, ie. the Atari Breakout game, there are 4 possible actions and the following table maps actions to corresponding integer values.

```{r, eval=TRUE, echo=FALSE}
knitr::kable(data.frame(
  "Action" = c("NOOP (Do nothing)", "FIRE (Get the ball)", "RIGHT", "LEFT" ),
  "Integer" = c(0, 1, 2, 3)
))
```

## Pre-processing

Now that we know about Gym, let's define some functions to preprocess the state before saving it to our experience memory. We will transform each frame to grayscale since the colors doesn't matter in this game and then we will rescale the image to save RAM memory.

We implemented 3 functions to do this:

```{r}
# transforms the input image to grayscale
grayscale <- function(x) {
  dims <- c(dim(x)[-3], 1)
  x <- (x[,,1, drop = FALSE] + x[,,2, drop = FALSE] + x[,,3, drop = FALSE])/3
  array(as.integer(x), dim = dims)
}

# reduces the scale of the image by skiping some pixels
downsample <- function(x, by = 2) {
  dims <- dim(x)
  x[seq(1, to = dims[1], by = by), seq(1, to = dims[2], by = by),]
}

# does both downsampling and grayscale.
preprocess <- function (x) {
  x %>%
    downsample(by = 4) %>%
    grayscale()
}
```

It's important to note that inside `grayscale` we transformed the resulting array to an integer array in order to save memmory.

## Experience

Now let's define an auxiliary class that will help us to keep the replay memory. Since we have to keep a large list of states, we need to take care to avoid copies. This is challeging in R because of the *copy-on-modify* beahvior.[@wickham2014advanced]. We implement an `R6` class manage storing and sampling from the replay memory.

```{r}
Experience <- R6::R6Class(
  "Experience",

  public = list(

    # class initialization
    initialize = function(max_size = 100000) {
      private$memory <- vector(mode = "list", max_size)
      private$max_size <- max_size
    },

    # adds transitions to the replay memory
    push = function(s_t, s_t1, terminal, action, reward) {

      private$memory[[private$i]] <- list(
        s_t = s_t,
        s_t1 = s_t1,
        terminal = as.logical(terminal),
        action = as.integer(action),
        reward = reward
      )

      if (private$i == private$max_size) {
        private$i <- 1L
        private$full <- TRUE
      } else {
        private$i <- private$i + 1 
      }

      invisible(TRUE)
    },

    # sample elements from the replay buffer
    sample = function(size = 32) {

      n <- ifelse(private$full, private$max_size, private$i - 1)
      ids <- sample.int(n, size)

      batch <- purrr::transpose(private$memory[ids])

      list(
        s_t = abind::abind(batch$s_t, along = 0.1),
        s_t1 = abind::abind(batch$s_t1, along = 0.1),
        terminal = unlist(batch$terminal),
        action = unlist(batch$action),
        reward = unlist(batch$reward)
      )
    }

  ),

  private = list(
    memory = NULL,
    i = 1,
    full = FALSE,
    max_size = NULL
  )

)
```

With this implementation we can create an instance of the `Experience` class with:

```{r}
experience <- Experience$new(max_size = 1000)
```

Then we can add transitions and sample from the replay buffer with:

```{r}
experience$push(s_t, s_t1, terminal, action, reward)
experience$sample(size = 32)
```

Note that the `experience` object is modified in place when pushing elements.
